---
title: "LMDC Misdeamenor Charges and Marijuana Charges"
author: "Jacob Mercer"
date: "5/8/2019"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(data.table)
library(knitr)

setwd('C:\\Users\\Jake\\Documents\\R\\EquityAnalysis\\2_ModFiles')

lmdc_raw <- fread('LMDC_pop.csv')
  
```

```{r data-prep}
lmdc <- lmdc_raw %>%
  mutate(
    BookingNumber = str_trim(BookingNumber, "both"),
    BookingDate = ymd_hms(BookingDate),
    SnapshotDateTime = ymd_hms(SnapshotDateTime),
    RaceDescription = str_trim(RaceDescription, 'both'),
    RaceDescription = if_else(RaceDescription == 'UNKNOWN', 'Other', RaceDescription)
  ) %>%
  filter(BookingDate >= '2014-07-11')


# pull out all BNs that have charges *other* than MISDEMEANOR

lmdc_not_mis <- lmdc %>% 
  filter(ChargeCategory != 'MISDEMEANOR') %>%
  distinct(BookingNumber)

# use lmdc_not_mis to get all BNs from LMDC that are only brought in on MISDEMEANOR charges

lmdc_mis <- lmdc %>%
  filter(!BookingNumber %in% lmdc_not_mis$BookingNumber)

lmdc_mis_BNs <- lmdc_mis %>%
  distinct(BookingNumber) %>%
  filter(!BookingNumber %in% lmdc_not_mis$BookingNumber)


# to look at what people are charged with when brought in, pull out all BNs that have the min snapshot datetime

mis_init_booking <- lmdc_mis %>%
  group_by(BookingNumber) %>%
  filter(SnapshotDateTime == min(SnapshotDateTime))
```


## Misdemeanor Charges
Filtered LMDC booking data (July 2014 through June 2016) to look at individuals who are intially booked exclusively on misdeamenor charges


```{r outline-data}
# Individuals
mis_init_booking_by_race <- mis_init_booking %>%
  group_by(BookingNumber, RaceDescription) %>%
  summarize() %>%
  group_by(RaceDescription) %>%
  summarize(Count = n()) %>%
  arrange(desc(Count))

kable(mis_init_booking_by_race, caption = "Racial Composition of Individuals Booked on Misdemeanor Charges")

mis_init_booking %>%
  mutate(
    BookingDate = as_date(BookingDate),
    year_mo = str_sub(BookingDate, end = 7)
  ) %>%
  group_by(year_mo, RaceDescription) %>%
  summarize(n = n(),
            daily_avg = n() / 30) %>%
  ggplot() +
  geom_col(aes(x = year_mo, y = daily_avg, fill = RaceDescription)) +
  labs(title = 'Monthly Average of Misdemeanor Bookings Per Day') +
  ylab('Count') +
  xlab('Year-Month') +
  theme(axis.text.x = element_text(size = 10, angle = 45))

# Age Breakdown
mis_init_booking %>%
  group_by(BookingNumber, CurrentAge, RaceDescription) %>%
  summarize() %>%
  ggplot() +
  geom_freqpoly(aes(x = CurrentAge, color = RaceDescription), binwidth = 2) +
  labs(title = 'Age Distribution by Race') +
  ylab('Count') +
  xlab('Age')

```

## Marijauna Charges

Further focused on marijuana misdemeanor offenses for two reasons

* Examine racial disparity in marijuana charges
* Get a sense for the effect of decriminalization or legalization

### Racial Disparities

```{r mj-charges}
# filter lmdc_mis to look at marijuana offences

mj_mis_BNs <- mis_init_booking %>%
  filter(str_detect(Charge, 'MARIJUANA')) %>%
  distinct(BookingNumber)

mj_mis <- mis_init_booking %>%
  filter(BookingNumber %in% mj_mis_BNs$BookingNumber)

# look at top charges by race

mj_mis_top_n <- mj_mis %>%
  group_by(Charge) %>%
  summarize(n = n()) %>%
  top_n(3, n)

mj_mis %>%
  filter(Charge %in% mj_mis_top_n$Charge) %>%
  mutate(
    RaceCode = ifelse(RaceCode == 'W', 'W', 'not W')
  ) %>%
  ggplot() +
  geom_bar(aes(x = Charge, fill = RaceCode), position = 'dodge') +
  coord_flip() +
  labs(title = 'Top 3 Misdemeanor Charges') +
  ylab('Count') +
  xlab('') +
  theme(axis.text.x = element_text(size = 10, angle = 45)) +
  theme(axis.text.y = element_text(size = 10))

mj_mis_w <- mj_mis %>%
  filter(RaceCode == 'W') %>%
  group_by(Charge) %>%
  summarize(Count = n()) %>%
  top_n(10, Count) %>%
  arrange(desc(Count))

kable(mj_mis_w, caption = "Top 10 Misdemeanor Charges for White Individuals")

mj_mis_w <- mj_mis %>%
  filter(RaceCode != 'W') %>%
  group_by(Charge) %>%
  summarize(Count = n()) %>%
  top_n(10, Count) %>%
  arrange(desc(Count))

kable(mj_mis_w, caption = "Top 10 Misdemeanor Charges for Non-White Individuals")


```


### Decriminalization

An examination of the effect if the state of Kentucky no longer required an individual must be bought into custody for marijuana offenses. 

For more information on the methdology, please see the folloiwng Note on Assumptions.

```{r mj-decrim-calc}

removed_charges <- mj_mis %>%
  ungroup() %>%
  filter(str_detect(Charge, 'DEADLY') 
         | str_detect(Charge, 'U/INFLUENCE') 
         | str_detect(Charge, 'ROBBERY')
         | str_detect(Charge, 'ASSAULT')
         | str_detect(Charge, 'ENHANCEMENT')
         | str_detect(Charge, 'EPO')) %>%
  distinct(Charge)

# get BNs that have these charges and use to filter mj_mis to get BNs that don't

removed_charges_BNs <- mj_mis %>%
  filter(Charge %in% removed_charges$Charge) %>%
  distinct(BookingNumber)

included_charges <- mj_mis %>%
  ungroup() %>%
  filter(!Charge %in% removed_charges$Charge) %>% 
  distinct(Charge)

mj_non_v_BNs <- mj_mis %>%
  filter(!BookingNumber %in% removed_charges_BNs$BookingNumber) %>%
  distinct(BookingNumber)

mj_non_v_offenders <- mj_mis %>% filter(BookingNumber %in% mj_non_v_BNs$BookingNumber) %>%
  group_by(BookingNumber, RaceDescription) %>%
  summarize() %>%
  group_by(RaceDescription) %>%
  summarize(Count = n()) %>%
  arrange(desc(Count))

mj_non_v_time_booked <- lmdc %>%
  filter(BookingNumber %in% mj_non_v_BNs$BookingNumber) %>%
  group_by(BookingNumber) %>%
  summarise(BookingDate = min(BookingDate),
            EndDate = max(SnapshotDateTime),
            TimeBooked = max(SnapshotDateTime) - min(BookingDate),
            EstCost = as.integer(TimeBooked) * 82)

total_cost <- sum(mj_non_v_time_booked$EstCost)
median_time_booked <- median(mj_non_v_time_booked$TimeBooked)
start_date <- min(mj_non_v_time_booked$BookingDate)
end_date <- max(mj_non_v_time_booked$EndDate)
cost_per_week <- round(total_cost / as.numeric(end_date - start_date), 2) * 7

```

```{r mj-decrim-show, echo=FALSE}
kable(mj_non_v_offenders, caption = 'Individuals who would not be booked')
```

The median time over which inidivuals are booked:

```{r median-time-booked, echo = FALSE}
median_time_booked
```

Estimating that it costs $82[^1] per individual per day to house booked individuals, this is the total cost of holding:

```{r total-cost, echo = FALSE}
total_cost
```

This is over the entire time period of the dataset:

```{r date-range, echo=F}
start_date
end_date
```

Which leads to an average weekly cost of:

```{r cost-per-week, echo=F}
cost_per_week
```



### Note on Assumptions

Since virtually all individuals booked have multiple charges, the following charges are inlcuded along with marijuana offenses. The working assumption for the above analysis is that these charges do *not* require an individual to be booked and that the marijuana charges were the precipitating factor. However given the subjective nature of police work and that each incident is unique, the above analysis should be taken as an estimation of general trends and not a rigorous statement of fact.

Generally, the misdemeanors removed from consideration are any robber, driving under the influence, possession of a deadly weapon, assault, or EPO violation.


```{r charged-included}
kable(included_charges, caption='Charges Included')
```

### References
[Data is drawn from Louisville Metro Dept of Corrections](https://data.louisvilleky.gov/dataset/daily-lmdc-population-snapshots) and covers Jefferson County, KY from the time frame from 2014-07-11 to 2016-06-29

[^1]: $82 per individual per day figure is from conversationa with City of Louisville's Center for Health Equity
