---
title: "Gender and Traffic Stops"
author: "Jacob Mercer"
date: "6/24/2019"
output: 
  html_document:
    code_folding: hide
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(data.table)
library(knitr)

stops_raw <- fread("C:/Users/Jake/Documents/Repos/Rstuff/LPD_stops/source_data/LMPD_STOPS_DATA_12.csv")

```

```{r data transform}

stops <- stops_officer_m <- stops_raw %>%
  filter(OFFICER_GENDER != '') %>%
  filter(`ACTIVITY RESULTS` != '') %>%
  filter(ACTIVITY_BEAT != '' & ACTIVITY_BEAT != ' ') %>%
  filter(DRIVER_GENDER != '') %>%
  filter(DRIVER_AGE_RANGE != "") %>%
  filter(DRIVER_RACE != "") %>%
  filter(`NUMBER OF PASSENGERS` != "") %>%
  filter(`NUMBER OF PASSENGERS` <= 6) %>%
  filter(WAS_VEHCILE_SEARCHED != "") %>%
  filter(DRIVER_RACE == 'BLACK'| DRIVER_RACE == 'HISPANIC' | DRIVER_RACE == 'WHITE') %>%
  filter(ACTIVITY_TIME != "" & ACTIVITY_TIME != " ") %>%
  filter(ACTIVITY_DIVISION != "NON LMPD" & ACTIVITY_DIVISION != "") %>%
  mutate(
    activity_date = mdy(ACTIVITY_DATE),
    date_time = ymd_hms(str_c(activity_date, ACTIVITY_TIME))
  ) %>%
  filter(activity_date < ymd(20160101)) %>%
  mutate(
    cat_month = month(date_time),
    cat_wday = wday(date_time),
    cat_mday = mday(date_time),
    cat_time = cut(hour(date_time), c(0, 4, 8, 12, 16, 20, 24))
  ) %>%
  filter(!is.na(cat_time))

stops_officer_m <- stops %>%
  filter(OFFICER_GENDER == 'M') %>%
  mutate(
    `Driver Gender` = DRIVER_GENDER,
    Outcome = `ACTIVITY RESULTS`
  ) %>%
  group_by(`Driver Gender`, Outcome) %>%
  summarize(N = n()) %>%
  mutate(
    Count = N / sum(N),
    Percentage = round((Count * 100), 0)
  )

stops_officer_f <- stops %>%
  filter(OFFICER_GENDER == 'F') %>%
  mutate(
    `Driver Gender` = DRIVER_GENDER,
    Outcome = `ACTIVITY RESULTS`
  ) %>%
  group_by(`Driver Gender`, Outcome) %>%
  summarize(N = n()) %>%
  mutate(
    Count = N / sum(N),
    Percentage = round((Count * 100), 0)
  )

officer_divisions <- stops %>%
  group_by(OFFICER_GENDER, ACTIVITY_DIVISION) %>%
  summarize(N = n()) %>%
  mutate(
    freq = N / sum(N),
    perc = round((freq * 100), 0)
  )

```

### Officer Gender and Traffic Stops

The Louisville Metro Police Department’s traffic stop data presents an opportunity to examine trends in police stops. The data reviewed here is from the year of 2015.

This analysis started with the question, “Are male police officers more likely than female police officers to give women drivers warnings as opposed to citations during traffic stops?”  This question examines the common trope that male officers are more forgiving of female drivers than male drivers.

The answer to this question revealed another trend in the data that warranted further exploration. 



```{r graph 1}
p <- ggplot(data = stops_officer_m, mapping = aes(x = `Driver Gender`, y = Percentage))
p + geom_col(aes(fill = Outcome), position = "dodge") +
  labs(x = "Gender", 
       y = 'Percentage',
       title = "Stops by Male Officers",
       subtitle = "Percentage of Warnings vs Citations given by gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Stop Outcome")
```

### Male Officer Stops

``` {r table 1}
kable(stops_officer_m)
```

```{r graph 1.2}
p <- ggplot(data = stops_officer_f, mapping = aes(x = `Driver Gender`, y = Percentage))
p + geom_col(aes(fill = Outcome), position = "dodge") +
  labs(x = "Gender", 
       y = 'Percentage',
       title = "Stops by Female Officers",
       subtitle = "Percentage of Warnings vs Citations given by gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Stop Outcome")
```

### Female Officer Stops

```{r table 2}
kable(stops_officer_f)
```


In 2015, it does not appear that female drivers received any particular deference from male officers. In fact, there does not appear to be any meaningful discrepancy between how officers of either gender treated male or female traffic offenders.

There is a marked difference in the ratio of citations to warnings issued, depending on the gender of the officer.

The LMPD traffic stops data set contains several variables that could be corollary or even explanatory with regard to the observed discprancy. That said, examining differences between male and female behavior is a dense question with confounding variables and causes that defy reliable measurement. It is important to keep these limitations in mind.

Diving into other parts of the data set, the age range and race of drivers who were stopped does not seem to show a marked difference. Additionally, the number of passengers in the vehicle at the time of the stop does not show much disparity between male and female officers.



```{r graph 2}
p <- ggplot(data = stops, mapping = aes(x = DRIVER_AGE_RANGE))
p + geom_bar(aes(fill = OFFICER_GENDER, y = ..prop.., group = OFFICER_GENDER), position = "dodge") +
  labs(x = "Age", 
       y = 'Percentage',
       title = "Age Range of Drivers Stopped by Officer Gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Officer Gender")


p <- ggplot(data = stops, mapping = aes(x = `NUMBER OF PASSENGERS`))
p + geom_bar(aes(fill = OFFICER_GENDER, y = ..prop.., group = OFFICER_GENDER), position = "dodge") +
  labs(title = "Number of Passengers by Officer Gender") +
  labs(x = "Number of Passengers", 
       y = 'Percentage',
       title = "Number of Passengers in Vehicle by Officer Gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Officer Gender")

p <- ggplot(data = stops, mapping = aes(x = DRIVER_RACE))
p + geom_bar(aes(fill = OFFICER_GENDER, y = ..prop.., group = OFFICER_GENDER), position = "dodge") +
  labs(title = "") +
  labs(x = "Race", 
       y = 'Percentage',
       title = "Driver Race by Officer Gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Officer Gender")

```


Looking at the dates and time of the stops does seem to show some differences.


```{r graph 3}
p <- ggplot(data = stops, mapping = aes(x = cat_month))
p + geom_bar(aes(fill = OFFICER_GENDER, y = ..prop.., group = OFFICER_GENDER), position = "dodge") +
  labs(x = "Month", 
       y = 'Percentage',
       title = "Month of Stop by Officer Gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Officer Gender") +
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7,8,9,10,11,12),
                     labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))


p <- ggplot(data = stops, mapping = aes(x = cat_mday))
p + geom_bar(aes(fill = OFFICER_GENDER, y = ..prop.., group = OFFICER_GENDER), position = "dodge") +
  labs(x = "Month", 
       y = 'Percentage',
       title = "Day of The Month of Stop by Officer Gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Officer Gender") +
  scale_x_continuous(breaks = c(1,5,10,15,20,25,30))

p <- ggplot(data = stops, mapping = aes(x = cat_wday))
p + geom_bar(aes(fill = OFFICER_GENDER, y = ..prop.., group = OFFICER_GENDER), position = "dodge") +
  labs(x = "Day", 
       y = 'Percentage',
       title = "Day of the Week of Stop by Officer Gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Officer Gender") +
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7),
                     labels = c("Sun", "Mon", "Tues", "Weds", "Thurs", "Fri", "Sat"))

p <- ggplot(data = stops, mapping = aes(x = cat_time))
p + geom_bar(aes(fill = OFFICER_GENDER, y = ..prop.., group = OFFICER_GENDER), position = "dodge") +
  labs(x = "Hour", 
       y = 'Percentage',
       title = "Time of Stop by Officer Gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Officer Gender") +
  scale_x_discrete(labels = c("0-4", "4-8", "8-12", "12-16", "16-20", "20-24"))


```


There are several trends that appear relevant, when comparing how male and female officers shifts are allocated.

* A given female officer is more likely to work weekends
* A given female officer is more likely to work between the hours of 8pm and 4 am

The starkest difference between male and female officers comes from the geographic area where the traffic stop occurred.



```{r graph 4}
p <- ggplot(data = officer_divisions, mapping = aes(x = OFFICER_GENDER, y = perc))
p + geom_col(aes(fill = ACTIVITY_DIVISION), position = "dodge2") +
  labs(x = "Gender", 
       y = 'Percentage',
       title = "Division Distribution by Officer Gender",
       caption = "Data: LMPD Stops Data (https://data.louisvilleky.gov/dataset/lmpd-stops-data)") +
  scale_fill_discrete(name = "Division")
```


#### Conclusions

There are three areas of divergence between male and female officers that could provide clues as to why female officers issue more warnings.

The first two are that a  female officer is more likely to work nights and weekends than a male officer. There are several ways that this could result in female officers giving more warnings, including stopping erratic but not impaired drivers and drivers with cars with broken lights. 

The third area is that female officers are more likely than their male counterparts to patrol the 8th Division. This is an area that covers the eastern Louisville suburbs, which are relatively wealthy. It is possible that a deference to more affluent drivers is influencing the decision to write fewer tickets.

There are also other possible explanations that this study is not equipped to address. For example, it could be that female officers pull over more people than their male counterparts, which could result in more warnings.

Generally speaking, there does not seem to be solid evidence that could point to why female officers are more generous with the drivers that they stop than male officers. All that can be known for certain is that the trend exists.


#### Further Study

*	Determine if female officers perform traffic stops at a significantly higher rate than male officers
*	Examine the characteristics of the different divisions and see if there are any useful correlations that can be drawn
*	Determine the character of traffic stops during the nights and weekends as opposed to those during the day
*	Directly poll male and female officers to gain insight into traffic stop habits


#### Citation
All data is from Louisville Metro Open Data. The LMPD Stops Data can be found here:
https://data.louisvilleky.gov/dataset/lmpd-stops-data

